---
title: "Data Pre-processing"
author: "Marly"
date: "2024-08-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load packages
```{r}
library(microbiome) # data analysis and visualisation
library(phyloseq) # also the basis of data object. Data analysis and visualisation
library(microbiomeutilities) # some utility tools
library(RColorBrewer) # nice color options
library(ggpubr) # publication quality figures, based on ggplot2
library(DT) # interactive tables in html and markdown
library(data.table) # alternative to data.frame
library(dplyr) # data handling
```


```{r}
#load the phylooseq objects

ps <- readRDS("../phyloseq/Namibia_samples_decont.rds")

ps
```


```{r}
data <- as.data.frame(tax_table(ps)) # the table is interactive you can scrol and search thorugh it for details.
```


# check for features of data  
```{r}
summarize_phyloseq(ps)

ps
```

sample_sums
```{r}
 ?sample_sums
p <- barplot(sample_sums(ps))
```

Variablity

We will check the spread of variation for taxa identifed. Coefficient of variation (C.V), i.e. sd(x)/mean(x) is a widely used approach to measure heterogeneity in OTU/ASV abundance data. The plot below shows CV-mean(relaive mean abundance) relationship in the scatter plot, where variation is calculated for each OTU/ASV across samples versus mean relative abundance.
Now plot C.V.

```{r}
# the plot_taxa_cv will first convert the counts to relative abundances and then calculate the C.V.
p1 <- plot_taxa_cv(ps, plot.type = "scatter")
```
```{r}
p1 + scale_x_log10()

## Scale for 'x' is already present. Adding another scale for 'x', which will
## replace the existing scale.
```

From the above two plots, we see that there are several OTUs which have high C.V. and low mean. 
Note:  The observed high variance shown here can also be explained in part by the difference between sample groups (larvae, water, controls) biology of samples is also important.

Let us check for distribution of number of sequences retained after OTU picking and NG-tax approach.
```{r}
p_seqdepth.otu <- plot_read_distribution(ps, "sample_or_control", "density")
print(p_seqdepth.otu)

tax_table <- as.data.frame(tax_table(ps))

```
Moving on to distribution of taxa
```{r}
# We make a data table with information on the OTUs
ps0_df_taxa <- data.table(tax_table(ps),
                          ASVabundance = taxa_sums(ps),
                          ASV = taxa_names(ps))
```

```{r}
ps0_tax_plot <- ggplot(ps0_df_taxa, aes(ASVabundance)) +
  geom_histogram() + ggtitle("Histogram of counts") +
  theme_bw() + scale_x_log10() + ylab("Frequency of Counts") + xlab("Abundance (raw counts)")

print(ps0_tax_plot)
```

The OTU distribution is not a normal distribution. This has implications in downstream statistical analysis like differential abundance testing.

We can also check for prevalance abundance distribtuion of ASVs and OTUs.

```{r}
p <- plot_taxa_prevalence(ps, "Phylum")
p
```

OTU-picking stratergy

```{r}
ps0.1.otu <- subset_taxa(ps, Phylum != "")

print(ps0.1.otu)
```

```{r}
p <- plot_taxa_prevalence(ps0.1.otu, "Phylum")
p
```

# also check how many are lost 
```{r}
ntaxa(ps)-ntaxa(ps0.1.otu)
```

#total number of reads in the dataset
```{r}
reads_per_otu <- taxa_sums(ps0.1.otu)
print(sum(reads_per_otu))
```

There are 5763961 reads in the total data set.
How many OTUs are less than 10 reads and how many reads do they contain?

```{r}
print(length(reads_per_otu[reads_per_otu < 10]))
```

```{r}
print(sum(reads_per_otu[reads_per_otu < 10]))
```
To put this into context; out of the 63611 OTUs, 308 OTUs contain less than 10 reads, which is:
```{r}
print((305/62461)*100)
```

Let us see how many singletons are there?
```{r}
length(which(taxa_sums(ps0.1.otu) <= 1))
```
Check how many doubletons are there?
```{r}
length(which(taxa_sums(ps0.1.otu) == 2))
```

Check how many Singletons and doubletons are there?
```{r}
length(which(taxa_sums(ps0.1.otu) <= 2))
```

Singletons and doubletons
```{r}
round((224/62461)*100, digits = 2)
```


3.6% of the OTUs are doubletons or singletons. This is suggests that there can be potentially spurious OTUs.



